---
title:  "10,000 Foot Overview of R"
author: "Andreas Santucci"
output: html_document
---

R is designed to facilitate statistical analysis. To that end, it makes
common data-operations such as ingestion, re-organization, visualization, and modeling quite simple to execute.

# Loading Data
We often want to start by loading some data into R. There are several basic ways. We may use the `read.table` command to read from a variety of different inputs. For example, the [following page](http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.info.txt) hosts a LA-ozone data-file.

```{r}
webSite <- 'http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.data'
data <- read.table(webSite, sep = ',', header = TRUE)
head(data)
```

The main argument to `read.table` is a string dictating where the data-file is to be read from, whether it be locally stored on disk or hosted online.

# Inspecting Data
What are the dimensions of our data?

```{r}
dim(data)
```

When our data is read, it is converted to a type of object in R called a `data.frame`. This is a way to hold a collection of vectors (numeric, character, or mixed) each the same length but possibly different types. We may extract a subset of rows and columns easily.

```{r}
data[3:6, 2:ncol(data)]
```

We might wish to know how our data are *structured*, in which case we may use the `str` command.

```{r}
str(data)
```

This tells us the dimensions of our data, the column names, their respective data-types, and the first few entries in each column. We might also apply the
`summary` function to learn several statistics of our data.

```{r}
summary(data)
```

# Visualizing Data
We can easily display graphical summaries of our data using basic commands like
`hist` and `plot` for histograms and two-way scatter plots respectively.

```{r}
plot(data[, 1:4])        # Plot pairs among first several variables.
hist(data$ozone)
```

# Modeling
We can easily construct a *linear model* using `lm`.

```{r}
model <- lm(vis ~ temp, data)  # Regress visibility as a function of temperature.
summary(model)
```

We can easily update an existing model; the formula language is very expressive.

```{r}
### Add humidity and the interaction of temperature with humidity.
update(model, . ~ . + humidity + temp*humidity)
```

# R is Object Oriented
Notice that when the `summary` function is called, its behavior depends on its
input argument type. This is because each *object* in R is associated with a *class*, each of which come with a set of corresponding *methods*.

```{r}
head(summary.lm)
head(summary.data.frame)
```

In the next module, we'll dive into the basics a bit further.


