---
title:  "10,000 Foot Overview of R"
author: "Andreas Santucci"
output: html_document
---

# Loading Data
We often want to start by loading some data into R. There are several basic ways. We may use the `read.table` command to read from a variety of different inputs. For example, we can read data from a web-page.

```{r}
data <- read.table('http://www.stats.ox.ac.uk/pub/datasets/csb/ch11b.dat')
head(data)
```

# Inspecting Data
What are the dimensions of our data?

```{r}
dim(data)
```

When our data is read, it is converted to a type of object in R called a `data.frame`. This is a way to hold a collection of vectors (numeric, character, or mixed) each the same length but possibly different types. We may index the rows and columns easily

```{r}
data[3:10, 2:ncol(data)]
```

We might wish to know how our data are *structured*, in which case we may use the `str` command.

```{r}
str(data)
```

This tells us the dimensions of our data, the column names, their respective data-types, and the first few entries in each column. We might also apply the
`summary` function to learn several statistics of our data.

```{r}
summary(data)
```

# Visualizing Data
We can easily display graphical summaries of our data using basic commands like
`hist` and `plot` for histograms and two-way scatter plots respectively.

```{r}
hist(data$V3)
plot(data$V3, data$V4)
```

# Modeling
We can easily construct a *linear model* using `lm`.

```{r}
model <- lm(V4 ~ V3, data)
summary(model)
```

# R is Object Oriented
Notice that when the `summary` function is called, its behavior depends on its
input argument type. This is because each *object* in R is associated with a *class*, each of which come with a set of corresponding *methods*.



That's a 10,000 foot overview of what a typical work-flow in R may look like.

In the next module, we'll dive into the basics 
a bit further.
