---
title:  "Basic Operations in R"
author: "Andreas Santucci"
output: html_document
---

# R as an extensible calculator
We evaluate a couple elementary mathematical expressions.

```{r, results='hold'}
2 * pi
sqrt(100)
```

Notice that `pi` is a built-in constant. There are a select
few other constants defined for each R session, such as months

```{r, results='hold'}
head(month.abb)
tail(month.name)
```

or the alphabet.

```{r, results='hold'}
head(letters)
tail(LETTERS)
```

Both of these are examples of *character-vectors*, i.e. a vector in which each element itself is a character-string.

R makes it easy to work with dates.

```{r, results='hold'}
x <- seq(from = as.Date('Jan 01, 2004', format = '%b %d, %Y'),
         to   = as.Date('2004-12-31'),
         by   = 'day')

table(months(x))
which(x == '2004-09-01')
length(x)                  # Leap year!
```

Almost anything you can think to do is either built in as a basic operation
into R, available as part of a package, or can be programmed yourself.

```{r, results='hold'}
Sys.Date()     # What day  is it?
Sys.time()     # What time is it?

as.Date('2016-12-31') - Sys.Date() # Number of days until New Years.
```

# Vectors

The months and alphabets printed above are actually vectors.
It is easy to create a vector of numbers in different ways. We present
the `:` "through" operator, and the functions `c` and
`seq` ("concatenate" and "sequence" respectively).

```{r, results = 'hold'}
1:10               # Integers 1 through 10.
c(1, 10, 100)      # A vector with integers 1, 10, and 100.
seq(1, 10, by = 3) # A sequence of numbers, from 1 to 10, by 3's.
```

## Assignment Operator
We've seen already that we may use the assignment operator `<-` in order to
declare variables. Note that R is dynamically typed, which means that variable
names are not statically fixed to a particular type or class of object.

```{r, results='hold'}
x <- sample(letters, size = 10)
class(x)
```

```{r, results='hold'}
x <- sample(1:10)
y <- sample(0:1, size = length(x), replace = T)

class(x)
```

## Vector Arithmetic
Often, functions in base R are *vectorized*. That is, a function may be applied to an entire vector at once. Examples:

```{r, results='hold'}
sqrt(1:4)
10^(1:4)
1:4 + seq(0, pi, length.out = 4)
```

## Recycling
If an operator (such as `+`, `-`, or `^`) is applied to two vectors of different
lengths, R implicitly performs recycling by repeating the shorter vector (possibly a partial fraction of times) until the length of the longer vector is reached.

This means things can be recycled *evenly*...

```{r, results='hold'}
3   + 1:6   # The number '3' can be thought of as a vector of length one.
0:1 + 1:6   # Add a 1 to every other number from 1 to 6.
```

... or not, in which case a *warning* is triggered.

```{r}
1:3 * 1:5  # Oops!
```

Note that a warning is *not* an error; this means that your code will continue to run.

# Booleans in R
We may also use `TRUE` or `FALSE` (or `1` and `0`) as Boolean values. These may be used in (non)-obvious ways.

```{r}
if (sample(month.abb[1:3], 1) == 'Jan') {
    print("Lucky!")
} else {
    print("Unlucky!") 
}

if (1) print("Anything non-zero is true")
if (0) print("Zeros are always evaluated as false")

if (!FALSE) print("We can negate a logical using the ! operator")
```

Realize that we can also apply Boolean *arithmetic*

```{r, results='hold'}
TRUE + FALSE
mean(c(TRUE, TRUE, FALSE))
```

