---
title:  "R Basics"
author: "Andreas Santucci"
output: html_document
---

# Manipulation of Data
R is a fancy calculator.

```{r, results='hold'}
2 * pi
sqrt(100)
```

Notice that `pi` is a built-in constant. There are a select
few other constants defined for each R session, such as months

```{r, results='hold'}
head(month.abb)
tail(month.name)
```

or the alphabet.

```{r, results='hold'}
head(letters)
tail(LETTERS)
```

R makes it easy to work with dates.

```{r, results='hold'}
x <- seq(from = as.Date('Jan 01, 2004', format = '%b %d, %Y'),
         to   = as.Date('2004-12-31'),
         by   = 'day')
table(months(x))

which(x == '2004-09-01')
length(x)                  # Leap year!
```

## Vectors
The months and alphabets printed above are actually vectors.
It is easy to create a vector of numbers in different ways. We present
the `:` "through" operator, and the functions `c` and
`seq` ("concatenate" and "sequence" respectively).

```{r, results = 'hold'}
1:10               # Integers 1 through 10.
c(1, 10, 100)      # A vector with integers 1, 10, and 100.
seq(1, 10, by = 3) # A sequence of numbers, from 1 to 10, by 3's.
```

## Assignment Operator
We've seen already that we may use the assignment operator `<-` in order to
declare variables. Note that R is dynamically typed.

```{r, results='hold'}
x <- sample(letters, size = 10)
class(x)
```

```{r, results='hold'}
x <- sample(1:10)
y <- sample(0:1, size = length(x), replace = T)

class(x)
class(y)
```

R let's user's easily create vectors of numbers. We may assign numbers and 
vectors to variables.


often, functions in base R are *vectorized*. That is, a function may be applied to an entire vector at once.

```{r, results='hold'}
x <- seq(from = 0, to = 2*pi, by = 0.01)
length(x)
head(x)
tail(x)
```

```{r}
plot(x, sin(x), type = 'l')
```


# Objects in R
R is an *object-oriented* programming language. There are *classes* in R which
support various types of *methods*. Every object has a class.

```{r}
class(data)
class(model)
```

When we fed `data` (of type `data-frame`) to `summary`, the function first
inspected the object, determined its class, then went and called
the `summary.data.frame` method. Correspondingly, with the object `model` of class linear model, the method for `summary.lm` was called.

** Creating Objects
