---
title:  "Basic Operations and Data-Types in R"
author: "Andreas Santucci"
output: html_document
---

# R as an extensible calculator
We evaluate a couple elementary mathematical expressions.

```{r, results='hold'}
2 * pi
sqrt(100)
```

Notice that `pi` is a built-in constant. There are a select
few other constants defined for each R session, such as months

```{r, results='hold'}
head(month.abb)
tail(month.name)
```

or the alphabet.

```{r, results='hold'}
head(letters)
tail(LETTERS)
```

Both of these are examples of *character-vectors*, i.e. a vector in which each element itself is a character-string.

R makes it easy to work with dates.

```{r, results='hold'}
x <- seq(from = as.Date('Jan 01, 2004', format = '%b %d, %Y'),
         to   = as.Date('2004-12-31'),
         by   = 'day')

table(months(x))
which(x == '2004-09-01')
length(x)                  # Leap year!
```

Almost anything you can think to do is either built in as a basic operation
into R, available as part of a package, or can be programmed yourself.

```{r, results='hold'}
Sys.Date()     # What day  is it?
Sys.time()     # What time is it?

as.Date('2016-12-31') - Sys.Date() # Number of days until New Years.
```

# Vectors

The months and alphabets printed above are actually vectors.
It is easy to create a vector of numbers in different ways. We present
the `:` "through" operator, and the functions `c` and
`seq` ("concatenate" and "sequence" respectively).

```{r, results = 'hold'}
1:10               # Integers 1 through 10.
c(1, 10, 100)      # A vector with integers 1, 10, and 100.
seq(1, 10, by = 3) # A sequence of numbers, from 1 to 10, by 3's.
```

## Assignment Operator
We've seen already that we may use the assignment operator `<-` in order to
declare variables. Note that R is dynamically typed, which means that variable
names are not statically fixed to a particular type or class of object.

```{r, results='hold'}
x <- sample(letters, size = 10)
class(x)
```

```{r, results='hold'}
x <- sample(1:10)
y <- sample(0:1, size = length(x), replace = T)

class(x)
```

## Vector Arithmetic
Often, functions in base R are *vectorized*. That is, a function may be applied to an entire vector at once. Examples:

```{r, results='hold'}
sqrt(1:4)
10^(1:4)
1:4 + seq(0, pi, length.out = 4)
```

## Recycling
If an operator (such as `+`, `-`, or `^`) is applied to two vectors of different
lengths, R implicitly performs recycling by repeating the shorter vector (possibly a partial fraction of times) until the length of the longer vector is reached.

This means things can be recycled *evenly*...

```{r, results='hold'}
3   + 1:6   # The number '3' can be thought of as a vector of length one.
0:1 + 1:6   # Add a 1 to every other number from 1 to 6.
```

... or not, in which case a *warning* is triggered.

```{r}
1:3 * 1:5  # Oops!
```

Note that a warning is *not* an error; this means that your code will continue to run.

# Booleans in R
We may also use `TRUE` or `FALSE` (or `1` and `0`) as Boolean values. These may be used in (non)-obvious ways.

```{r}
if (sample(month.abb[1:3], 1) == 'Jan') {
    print("Lucky!")
} else {
    print("Unlucky!") 
}

if (1) print("Anything non-zero is true")
if (0) print("Zeros are always evaluated as false")

if (!FALSE) print("We can negate a logical using the ! operator")
```

Realize that we can also apply Boolean *arithmetic*

```{r, results='hold'}
TRUE + FALSE
mean(c(TRUE, TRUE, FALSE))
```
# Matrices
You've already seen vectors in R. Indexing matrices works the same way.

```{r}
m <- matrix(1:9, nrow = 3)
m[1:2, 1:2]
```

It turns out that actually, a matrix is simply a vector *with attributes*;
specifically, a matrix is a vector arranged by column.

```{r}
str(m)
m[4]  # Non-conventional: Access the fourth element in the matrix
```

Note that the entries in a matrix must be homogeneous. E.g., it is permissible
to have a matrix of numbers (or of characters), but not a matrix consisting of
some numbers and some characters.

```{r}
matrix(letters[1:9], nrow = 3)
```

# Lists
Lists are the most important construct in R. A list is a collection of named
(possibly heterogeneous) objects.

```{r}
myInfo <- list(first = "Andreas", last = "Santucci")

myInfo
```

Individual elements may be accessed from (or added to) a list by using either
the `$` *or* the `[[` operator. First we show the `$` operator for 
list-extraction.

```{r, results='hold'}
myInfo$first
myInfo$last
```

Now, we show the `[[` bracket operator in *two-different* ways; one way allows
us to access elements of a list by *name*, the other by *position*.

```{r, results='hold'}
myInfo[['first']]
myInfo[[1]]
identical(myInfo[['last']], myInfo[[2]])
```

With a list, we can really store a lot of different information.

```{r}
myInfo$weightHistory <- seq(0, 180, 10)
myInfo$Age <- 26
myInfo$siblings <- c(brothers = 1, sisters = 1)

myInfo
```

# Data-Frames
A data-frame is actually just a list of vectors, each of the same length.
```{r, results='hold'}
n <- 6
data <- data.frame(id = 1:n,
                   weight = runif(n, min = 100, max = 200),
                   birthm = sample(month.abb, size = n, replace = T))
                   
data[['id']]
data$weight
data[[ncol(data)]]
```

## Factors
Alongside `integer` and `character`, R also has a class of type `factor`. This is a class which posses *named levels*, each paired with a unique integer value.
Factors can be useful for regression modeling and plotting. Specifically,
they can model fixed-effects regressions with indicators very easily and also enable plotting by group in `ggplot2`.

However, there are some peculiarities to factors that may lead new users to be wary.

### R defaults to strings as factors
Notice that R default behavior is to read a string-column in as a factor variable in data-frames.

```{r}
head(data.frame)
default.stringsAsFactors()
```

### Factor labels are not the same as levels
One of the benefits of factors relies on how R stores them. However, 
it may feel awkward to new users that creating a factor of month-labels
yields levels which are placed in alphabetical (and not temporal)
order.

```{r}
as.factor(month.abb)
```

