---
title:  "Basic Operations and Data-Types in R"
author: "Andreas Santucci"
output: html_document
bibliography: bibliography.bib
nocite: | 
  @torfs_intro, @stack_resources, @r_inferno, @eddels_rev
---

# R as an extensible calculator
We evaluate a couple elementary mathematical expressions.

```{r}
2 * pi + sqrt(100)
```

Notice that `pi` is a built-in __constant__, and that `sqrt()` is a built-in
__function__.

### Creating a vector of numbers
We can easily create a vector of numbers and print the result
to console.

```{r}
1:40
```

### Interpreting the output
The `[1]` tells us the index of the first item in the row is 1, and if the output wraps around to a new-line, the subsequent `[.]`'s inform the index
for the following entry.

We may also feeding the anonymous vector as an input argument to another function.

```{r}
sum(1:100)
```

## Built-in Constants
There are *countless other functions* in R but only a *select few other constants* defined. One such example is **months**

```{r, results='hold'}
head(month.abb)
tail(month.name)
```

or the **alphabet**.

```{r, results='hold'}
head(letters)
tail(LETTERS)
```

## Characters and Strings in R
Both of these are examples of *character-vectors*, i.e. a vector in which each element itself is a character-string. In C, a character means a single byte (eight bits) which can represent unsigned integers 0-255, and a string is a sequence of null-terminated characters. In R, there is no notion of a character; the statistician is encouraged to think at the higher level of strings.

## Accessing individual elements of an array
We can access individual elements of a vector by using the `[` operator.

```{r}
letters[2]  # Emits second letter of the alphabet.
```

We can omit elements at specific indices using the `-` as a prefix operator.

```{r}
letters[-(1:20)]  # Omit first twenty letters of alphabet.
```

Of course, arrays are easy to work with. We can access their total length

```{r}
length(letters)
```

or even concatenate arrays using the `c` operator.

```{r}
c(letters, month.abb)
```

## Dates
R makes it easy to work with dates.

```{r, results='hold'}
x <- seq(from = as.Date('Jan 01, 2004', format = '%b %d, %Y'),
         to   = as.Date('2004-12-31'),
         by   = 'day')

table(months(x))
which(x == '2004-09-01')
length(x)                  # Leap year!
```

Almost anything you can think to do is either built in as a basic operation
into R, available as part of a package, or can be programmed yourself.

```{r, results='hold'}
Sys.Date()     # What day  is it?
Sys.time()     # What time is it?

as.Date('2016-12-31') - Sys.Date() # Number of days until New Years.
```


# Practice problem - Birthday countdown.
Ask the neighbor next to you their birthday.
Using R, generate a variable to hold your birthday and another
variable to hold your neighbors birthday.

Then, calculate the difference in age, measured in days.
```{r,echo=F,results='hide'}
myBirthday <- as.Date('1989-09-04')
myNeighbor <- as.Date('1995-12-31')
myNeighbor - myBirthday
```

# Vectors

The months and alphabets printed above are actually vectors.
It is easy to create a vector of numbers in different ways. We present
the `:` "through" operator, and the functions `c` and
`seq` ("concatenate" and "sequence" respectively).

```{r, results = 'hold'}
1:10               # Integers 1 through 10.
c(1, 10, 100)      # A vector with integers 1, 10, and 100.
seq(1, 10, by = 3) # A sequence of numbers, from 1 to 10, by 3's.
```

## Accessing elements of a Vector
All of the resulting arrays may have individual elements accessed,
either one at a time, all at once, or particular subsets.

```{r, results='hold'}
b <- 101:129
b[1:6]  # Access first six elements of 'b' vector.

idx <- c(1, 6, 14)
b[idx]      # Access only select elements in our index.
b[rev(idx)] # Access elements at positions 'idx' in *reverse* order.
b[-idx]     # Access every element in 'b' except those at 'idx' positions.
```

# Assignment Operator
We've seen already that we may use the assignment operator `<-` in order to
declare variables. Note that R is dynamically typed, which means that variable
names are not statically fixed to a particular type or class of object.

```{r, results='hold'}
z <- sample(letters, size = 10)
class(z)
```

```{r, results='hold'}
z <- sample(1:10)
y <- sample(0:1, size = length(x), replace = T)

class(x)
```

# Vector Arithmetic
Often, functions in base R are *vectorized*. That is, a function may be applied to an entire vector at once. Examples:

```{r, results='hold'}
sqrt(1:4)
10^(1:4)
1:4 + seq(0, pi, length.out = 4)
```

# Recycling
If an operator (such as `+`, `-`, or `^`) is applied to two vectors of different
lengths, R implicitly performs recycling by repeating the shorter vector (possibly a partial fraction of times) until the length of the longer vector is reached.

This means things can be recycled *evenly*...

```{r, results='hold'}
3   + 1:6   # The number '3' can be thought of as a vector of length one.
0:1 + 1:6   # Add a 1 to every other number from 1 to 6.
```

... or not, in which case a *warning* is triggered.

```{r}
1:3 * 1:5  # Oops!
```

Note that a warning is *not* an error; this means that your code will continue to run.

# Booleans in R
We may also use `TRUE` or `FALSE` (or `1` and `0`) as Boolean values. 
We may use Boolean vectors to indicate which elements to return when
slicing arrays.

# Practice problem - Array access
Create an expression which enables every element to be pulled out of object `b` which is divisible by 3.


```{r,echo=F,results='hide'}
head(b)
head(b %% 3)
head(b %% 3 == 0)
b[b %% 3 == 0]    # Get entries of 'b' which are a multiple of 3.
```

### Booleans and numbers
The fact that numeric values also evaluate to `TRUE` if non-zero, and `FALSE` is zero may be used in (non)-obvious ways. Conversely, `TRUE` and `FALSE` evaluate to zero and one respectively when used in mathematical expressions.

```{r}
if (1)   print("Anything non-zero is true")
if (-.4) print("Even really small negative values")
if (0)   print("Zeros are always evaluated as false")

if (!FALSE) print("We can negate a logical using the ! operator")
```

Realize that we can also apply Boolean *arithmetic*

```{r, results='hold'}
TRUE + FALSE
mean(c(TRUE, TRUE, FALSE))
```
# Matrices
You've already seen vectors in R. Indexing matrices works the same way.

```{r}
m <- matrix(1:9, nrow = 3)
m[1:2, 1:2]
```

To access all rows (or columns) just omit the corresponding index.

```{r}
m[1:2, ]  # Extract first two rows, and all columns.
```

### Relating Vectors and Matrices
It turns out that actually, a matrix is simply a vector *with attributes*;
specifically, a matrix is a vector arranged by column.

```{r}
str(m)
m[4]  # Non-conventional: Access the fourth element in the matrix
```

This is useful in some contexts, e.g. when we wish to apply a function
to all values in the matrix treated as a single input.

```{r}
mean(m)
```

### Note: Matrix entries must be homogeneous
Note that the entries in a matrix must be homogeneous. E.g., it is permissible
to have a matrix of numbers (or of characters), but not a matrix consisting of
some numbers and some characters.

```{r}
matrix(letters[1:9], nrow = 3)
```


# Lists
Lists are the most important construct in R. A list is a collection of named
(possibly heterogeneous) objects.

```{r}
myInfo <- list(first = "Joe", last = "Schmoe")

myInfo
```

Individual elements may be accessed from (or added to) a list by using either
the `$` *or* the `[[` operator. First we show the `$` operator for 
list-extraction.

```{r, results='hold'}
myInfo$first
myInfo$last
```

Now, we show the `[[` bracket operator in *two-different* ways; one way allows
us to access elements of a list by *name*, the other by *position*.

```{r, results='hold'}
myInfo[['first']]
myInfo[[1]]
identical(myInfo[['last']], myInfo[[2]])
```

With a list, we can really store a lot of different information.

```{r}
myInfo$weightHistory <- seq(0, 180, 10)
myInfo$Age <- 26
myInfo$siblings <- c(brothers = 1, sisters = 1)

myInfo
names(myInfo)
str(myInfo)
```

# Data-Frames
A data-frame is actually just a list of vectors, each of the same length.
```{r, results='hold'}
n <- 6
data <- data.frame(id = 1:n,
                   weight = runif(n, min = 100, max = 200),
                   birthm = sample(month.abb, size = n, replace = T))
                   
data[['id']]
data$weight
data[[ncol(data)]]
```

## A note on the `factor` class
Alongside `integer` and `character`, R also has a class of type `factor`. This is a class which posses *named levels*, each paired with a unique integer value.
Factors can be useful for regression modeling and plotting. Specifically,
they can model fixed-effects regressions with indicators very easily and also enable plotting by group in `ggplot2`.

However, there are some peculiarities to factors that may lead new users to be wary.

### R defaults to strings as factors
Notice that R default behavior is to read a string-column in as a factor variable in data-frames.

```{r}
head(data.frame)
default.stringsAsFactors()
```

A new user may find themselves always converting factors to strings, and subsequently may attempt to tamper the default settings in R. This is not recommended, since it means that your R code is no longer portable from one machine to another. For more, see the following StackOverflow [thread](http://stackoverflow.com/questions/8177921/how-to-disable-stringsasfactors-true-in-data-frame-permanently).

### Factor labels are not the same as levels
One of the benefits of factors relies on how R stores them. However, 
it may feel awkward to new users that creating a factor of month-labels
yields levels which are placed in alphabetical (and not temporal)
order.

```{r}
as.factor(month.abb)
```

Remember when we printed out the days in each month earlier?
```{r}
table(months(x))
```

Suppose we want to print the table in chronological (and not lexicographic) order.

```{r}
days <- data.frame(table(months(x)))
names(days) <- c('month', 'nDays')
days$month  <- as.character(days$month)

str(days)
```

```{r}
days$month <- factor(days$month, levels = month.name)
```
Specifically, the `factor` function takes an argument `levels`, for which we provide as value a character vector of month names in chronological order.

Now, when we request the R re-orders our data-frame according to the sorting
of the `month` variable, it will sort based on its `factor` ordering.

```{r}
days[order(days$month), ]
```

So, it turns out factors really are useful!

# Practice problem - using a factor in regression.
Obtain data from the [following page](https://web.stanford.edu/~hastie/ElemStatLearn/datasets/bone.data).
Regress \texttt{spnbmd} as a function of age and gender, using gender as a factor variable.
Compare this with creating an indicator variable for each of the $k-1$ levels in a $k$-level factor variable.

```{r, echo=F,results='hide'}
data <- read.table('https://web.stanford.edu/~hastie/ElemStatLearn/datasets/bone.data', header = T)
lm(spnbmd ~ age + gender, data)
```

# Resources for basic R
We cite "A (very) short introduction to R" by Torfs and Brauer, and also 
mention a [Stack Overflow Thread](http://stats.stackexchange.com/questions/138/free-resources-for-learning-r)
pointing to several great resources for learning R basics. 

For more detailed introductions, "The R Inferno" by Burns is often recommended. Personally, I learned many of my basics  from Joseph Adler's "R in a Nutshell".

# Bibliography