---
title:  "Functions in R"
author: "Andreas Santucci"
output: html_document
---

# Functions as objects
We can inspect the source-code for many R functions by simply typing the name of the function. For example, suppose we want to see how R implements a linear model.

```{r, results='hold'}
head(lm)
tail(lm)
```

How can we interpret what is being displayed? 
The function definition of `lm` is long, so we used `head` and `tail` to print out only the first few lines. 

In this printout we see two components: the function *arguments* and the beginning of the function *body*.

  * The function *arguments* of `lm` are named as `formula`, `data`, `subset`, `weights`,  etc.
  * The function *body* or definition begins after the `{` brace on line 4. 
  * The function *output* is the last expression in the body, just before the closing brace `}`.

It's interesting to note that the fact that we can even pass `lm` as an input argument to functions `head` and `tail` mean that *arguments to functions
may be functions themselves*.

# Defining Functions
We can define our own functions in R very easily.

```{r}
### Function: myFunction
  # --------------------
  # Input:  theta, measured in radians from 0 to 2*pi.
  # Output: A pair of points (x,y) representing the point on the unit circle 
  #         intersected by the input ray.
myFunction <- function(theta) {
    x <- cos(theta)
    y <- sin(theta)
    c(x, y)
}
```

We can immediately use our function.

```{r, results='hold'}
myFunction(0)
myFunction(pi/2)
myFunction(pi)
```



