---
title:  "Functions in R"
author: "Andreas Santucci"
output: html_document
---

# Functions as objects
We can inspect the source-code for many R functions by simply typing the name of the function. For example, suppose we want to see how R implements a linear model.

```{r, results='hold'}
head(lm)
tail(lm)
```

How can we interpret what is being displayed? 
The function definition of `lm` is long, so we used `head` and `tail` to print out only the first few lines. 

In this printout we see two components: the function *arguments* and the beginning of the function *body*.

  * The function *arguments* of `lm` are named as `formula`, `data`, `subset`, `weights`,  etc.
  * The function *body* or definition begins after the `{` brace on line 4. 
  * The function *output* is the last expression in the body, just before the closing brace `}`.

It's interesting to note that the fact that we can even pass `lm` as an input argument to functions `head` and `tail` mean that *arguments to functions
may be functions themselves*.

# Defining Functions
We can define our own functions in R very easily.

```{r}
### Function: myFunction
  # --------------------
  # Input:  theta, measured in radians from 0 to 2*pi.
  # Output: A pair of points (x,y) representing the point on the unit circle 
  #         intersected by ray emanating from the origin at angle theta.
myFunction <- function(theta) {
    x <- cos(theta)
    y <- sin(theta)
    c(x, y)
}
```

We can immediately use our function.

```{r, results='hold'}
myFunction(0)
myFunction(pi/2)
myFunction(pi)
```

# Modes of Calling functions
We defined a named function above. Some functions may actually come in the form of *binary operators*. For example, the addition operator.

```{r,results='hold'}
1 + 10     # Infix notation.
`+`(1, 10) # Prefix notation.
```

Notice that we may either use infix or prefix notation to call the function `+` on the arguments `1` and `10`.

# Functions as arguments. A Toy Example
Realize that in R, many functions are already vectorized, such as `+`.
```{r}
1:10 + 21:30
```

However, what if we were to consider how one might implement the `sum` operator given a higher order function `Reduce`. Reduce takes as argument a *binary-function* `f` and an object `x`. The function is applied on the elements of the input object `x`, and the result returned. 

```{r}
mySum  <- function(x) Reduce(`+`, x)
myProd <- function(x) Reduce(`*`, x)

identical( mySum(1:10),  sum(1:10))
all.equal(myProd(1:10), prod(1:10))  # Use all.equal to ignore type-difference.
```

# Next: `Apply` Functions
In the next section, we'll learn about `apply` functions, which themselves take in an object `X` with positive length and a function `FUN`; the function `FUN` is then applied to each element of object `X`.
